Predicting Success of Motivational Interviews with Recurrent Neural Network and Probabilistic Model for Patient-Provider Communication Sequences

### 2017-09-09 paper-amia-summit-2018/amia_summit_2018.tex
update word embedding part of the paper

\noindent{\bf Abstract}
\textit{The problem of analyzing temporally ordered observation sequences to make predictions related to the outcome of the processes that generated these sequences arises in many domains of healthcare informatics. In this paper, we focus on patient-provider communication sequences in the context of clinical interviews and propose Recurrent Neural Network (RNN) with two baseline methods Markov chain and Hidden Markov Model (HMM), for predicting the likelihood of eliciting a particular type of patient behavioral response based on an observed sequence of patient-provider exchanges. Our method achieved 70.03\%, 48.65\%, 88.53\% F1-score for under-sampled sequences and 78.96\%, 77.06\%, 87.16\% F1-score for over-sampled sequences in predicting the outcome of motivational interviews with obese adolescents using Markov Chain, HMM, and RNN, respectively. The proposed method can be used to automatically identify the most effective communication strategies in motivational interviews, which significantly decreases the effort required to develop effective interventions to address many public health conditions.}

@@

\section*{Methods}
\subsection*{\textit{Data collection}}
The experimental dataset for this work was constructed from the transcripts of 106 motivational interviews, which include a total of 50,239 segmented and annotated utterances. Each transcript consists of a counselor-adolescent and a counselor-caregiver session. Since the ultimate goal of a motivational interview is to increase the desire and ability of adolescents for a targeted behavior change, we considered only communication sequences from counselor-adolescent sessions and disregarded all communication sequences from counselor-caregiver sessions. The utterances were annotated based on MYSCOPE codebook \cite{carcone2013provider}, which are grouped into the adolescent and counselor code groups. Utterances were divided into successful and unsuccessful communication sequences. Successful communication sequences result in positive change talk and commitment language (a special class of change talk) statements by an adolescent, while unsuccessful sequences are the ones that result in negative change talk or commitment language and the sequences, in which no change talk or commitment language statements occur. Out of 4169 observed sequences, 4225 were positive and 918 were negative. For each of the probabilistic models (Markov chain and HMM), one model was trained using successful sequences and one model was trained using unsuccessful sequences. Statistics of the experimental dataset are presented in Table~\ref{tab:data_dist}. \\

@@

\textbf {MYSCOPE Code Embedding}: is a type of Neural Network, which can send feedback from the current hidden state to the hidden state of the next time step. In this way, RNN can capture long-term dependencies for predicting the future events, which is the main advantage of RNN over feedforward Neural Network, MC, and HMM. RNN can capture long-term dependencies for predicting the future events, which is the main advantage of RNN over feedforward Neural Network, MC, and HMM. Stochastic Neighbor Embedding for MYSCOPE code is shown in Figure~\ref{fig:code_embedding}.
\begin{figure}[!htb]
    \centering
    \includegraphics[width=0.50\textwidth]{figures/code_embed.eps}
    \caption{\textbf{T-SNE diagram of behavior code drawn after 4000 steps with perplexity 30}.}
    \label{fig:code_embedding}
\end{figure}

>>>>

\noindent{\bf Abstract}
\textit{The problem of analyzing temporally ordered observation sequences to make predictions related to the outcome of the processes that generated these sequences arises in many domains of healthcare informatics. In this paper, we focus on patient-provider communication sequences in the context of clinical interviews and propose Recurrent Neural Network (RNN) with two baseline methods Markov chain and Hidden Markov Model (HMM), for predicting the likelihood of eliciting a particular type of patient behavioral response based on an observed sequence of patient-provider exchanges. Our method achieved 70.38\%, 60.67\%, 86.77\% F1-score for under-sampled sequences and 77.75\%, 75.20\%, 83.81\% F1-score for over-sampled sequences in predicting the outcome of motivational interviews with obese adolescents using Markov Chain, HMM, and RNN, respectively. The proposed method can be used to automatically identify the most effective communication strategies in motivational interviews, which significantly decreases the effort required to develop effective interventions to address many public health conditions.}

@@

\section*{Methods}
\subsection*{\textit{Data collection}}
The experimental dataset for this work was constructed from the transcripts of 129 motivational interviews, which include a total of 50,239 segmented and annotated utterances. Each transcript consists of a counselor-adolescent and a counselor-caregiver session. Since the ultimate goal of a motivational interview is to increase the desire and ability of adolescents for a targeted behavior change, we considered only communication sequences from counselor-adolescent sessions and disregarded all communication sequences from counselor-caregiver sessions. The utterances were annotated based on MYSCOPE codebook \cite{carcone2013provider}, which are grouped into the adolescent and counselor code groups. Utterances were divided into successful and unsuccessful communication sequences. Successful communication sequences result in positive change talk and commitment language (a special class of change talk) statements by an adolescent, while unsuccessful sequences are the ones that result in negative change talk or commitment language and the sequences, in which no change talk or commitment language statements occur. Out of 5143 observed sequences, 4225 were positive and 918 were negative. For each of the probabilistic models (Markov chain and HMM), one model was trained using successful sequences and one model was trained using unsuccessful sequences. Statistics of the experimental dataset are presented in Table~\ref{tab:data_dist}. \\

@@

\textbf {MYSCOPE code embedding}: We utilize word embedding in place of one-hot vectors for the representation of behavior code. Because one-hot vectors are high-dimensional and sparse. On the other hand, word embeddings have the ability to represent semantically similar words with similar vectors in a lower dimensional space by retaining their internal relations. Figure~\ref{fig:code_embedding} illustrates the myscope code embedding visualize with t-SNE, a sophisticated technique for visualizing high-dimensional data. It can be seen that similar behavior codes are placed close to each other in high dimensional vector space such as OQECHT+, OQECML+, AF, ADV+, RCML+S, etc are nearest neighbors of CQECML+, which make sense in real world practice.   
\begin{figure}[!htb]
    \centering
    \includegraphics[width=0.50\textwidth]{figures/code_embed.eps}
    \caption{\textbf{T-SNE diagram of behavior code drawn after 4000 steps with perplexity 30}.}
    \label{fig:code_embedding}
\end{figure}
